---
title: "Test Tautges Bounds"
author: "Bret Larget"
date: "10/19/2017"
output: pdf_document
---

```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
```

Here is a short simulation to evaluate the bounds you found.
Not surprisingly,
they are not very tight.
My previous study indicated that the approximation should get closer
when the minimum alpha gets large,
yet the bounds you found are invariant to proportional changes to the alpha parameters.

Here I show what happens with $k=4$ as typical alphas vary from 2 to 20 to 200 to 2000 with the ratio of maximum lambda to minimum lambda being 10 and 1.1.

## Set-up

$Y_1,\ldots,Y_k \sim \text{independent Gamma}(\alpha_i,\lambda_i)$  
$S = \sum_{i=1}^k Y_i$  
$X_i = Y_i / S$ for $i = 1,\ldots,k$  
Let $m_i(\alpha,\lambda) = \mathsf{E}(X_i)$.  
Let $A_i(\alpha,\lambda) = (\alpha_i/\lambda_i) / \sum_{j=1}^k (\alpha_j/\lambda_j)$  
Let $\alpha_+ = \sum_{i=1}^k \alpha_i$. 
Let $\lambda_m = \min_i \lambda_i$ and $\lambda_M = \max_i \lambda_i$.

Tautges Bound:
$$
|m_i(\alpha,\lambda) - A_i(\alpha,\lambda)| \le \frac{\alpha_i(\lambda_M - \lambda_m)}{\alpha_+ \lambda_i})
$$

Note that for fixed $\lambda$ but $\alpha$ increasing proportionally, the bounds do not change.
I suspect that the true difference gets very small as $\alpha$ increases.
Test by simulation.
Pick random $\alpha$ and $\lambda$ with small,
but not tiny values.
Keep $\lambda$ fixed and see how $|m_i(\alpha,\lambda) - A_i(\alpha,\lambda)|$ changes when $\alpha$ is repeatedly multiplied by 10.

```{r}
## Pick parameters at random
parameters = function(k=4,mean.a=2000,shape=2,lambda.ratio=NULL)
{
  alpha = rgamma(k,shape,shape/mean.a)
  lambda = rgamma(k,shape,shape)
  if ( !is.null(lambda.ratio) )
  {
    if ( lambda.ratio < 1 )
      stop("lambda.ratio must be greater than or equal to one")
    if ( lambda.ratio == 1 )
      lambda = rep(1,k)
    else
      lambda = lambda + (max(lambda) - lambda.ratio*min(lambda))/(lambda.ratio-1)
  }
  lambda = lambda/mean(lambda)
  return(data.frame(alpha,lambda))
}

## Generate a large sample in order to
## numerically estimate the means
generate = function(p,n=1000)
{
  k = nrow(p)
  x = with(p, matrix(rgamma(n*k,rep(alpha,each=n),rep(lambda,each=n)),n,k) )
  s = apply(x,1,sum)
  for ( i in 1:n )
    x[i,] = x[i,] / s[i]
  return(x)
}

## Function to calculate bounds from p
bounds.tautges = function(p)
{
  lambda.max = max(p$lambda)
  lambda.min = min(p$lambda)
  out = with(p, alpha * (lambda.max - lambda.min) / (sum(alpha) * lambda))
  return ( out )
}

## approximate mean
approximate = function(p)
{
  w = with(p, alpha/lambda )
  return( w / sum(w) )
}

## Case where max(lambda) / min(lambda) = 10
p.0 = parameters(k=4,mean.a=2,lambda.ratio=10)

p.1 = with(p.0, data.frame(alpha=10*alpha,lambda=lambda))
p.2 = with(p.0, data.frame(alpha=100*alpha,lambda=lambda))
p.3 = with(p.0, data.frame(alpha=1000*alpha,lambda=lambda))

m.0 = apply(generate(p.0,n=1000000),2,mean)
m.1 = apply(generate(p.1,n=1000000),2,mean)
m.2 = apply(generate(p.2,n=1000000),2,mean)
m.3 = apply(generate(p.3,n=1000000),2,mean)

b.0 = bounds.tautges(p.0)
b.1 = bounds.tautges(p.1)
b.2 = bounds.tautges(p.2)
b.3 = bounds.tautges(p.3)

## approximation is the same regardless of mutiple of alpha
a.0 = approximate(p.0)

## difference
d.0 = m.0 - a.0
d.1 = m.1 - a.0
d.2 = m.2 - a.0
d.3 = m.3 - a.0

## results
df.0 = data.frame(p.0,mean=m.0,approx=a.0,diff=d.0,bound=b.0)
df.1 = data.frame(p.1,mean=m.1,approx=a.0,diff=d.1,bound=b.1)
df.2 = data.frame(p.2,mean=m.2,approx=a.0,diff=d.2,bound=b.2)
df.3 = data.frame(p.3,mean=m.3,approx=a.0,diff=d.3,bound=b.3)

## show results
show.results = function(df)
{
  cat("Lambda Ratio =",max(df$lambda)/min(df$lambda),"\n")
  cat("\n","alpha and lambda","\n",sep="")
  print(t(select(df,alpha,lambda)))
  cat("\n","means and approximations","\n",sep="")
  print(t(select(df,mean,approx)))
  cat("\n","differences and bounds","\n",sep="")
  print(t(select(df,diff,bound)))
}

show.results(df.0)
show.results(df.1)
show.results(df.2)
show.results(df.3)

## Case where max(lambda) / min(lambda) = 1.1
p.0 = parameters(k=4,mean.a=2,lambda.ratio=1.1)

p.1 = with(p.0, data.frame(alpha=10*alpha,lambda=lambda))
p.2 = with(p.0, data.frame(alpha=100*alpha,lambda=lambda))
p.3 = with(p.0, data.frame(alpha=1000*alpha,lambda=lambda))

m.0 = apply(generate(p.0,n=1000000),2,mean)
m.1 = apply(generate(p.1,n=1000000),2,mean)
m.2 = apply(generate(p.2,n=1000000),2,mean)
m.3 = apply(generate(p.3,n=1000000),2,mean)

b.0 = bounds.tautges(p.0)
b.1 = bounds.tautges(p.1)
b.2 = bounds.tautges(p.2)
b.3 = bounds.tautges(p.3)

## approximation is the same regardless of mutiple of alpha
a.0 = approximate(p.0)

## difference
d.0 = m.0 - a.0
d.1 = m.1 - a.0
d.2 = m.2 - a.0
d.3 = m.3 - a.0

## results
df.0 = data.frame(p.0,mean=m.0,approx=a.0,diff=d.0,bound=b.0)
df.1 = data.frame(p.1,mean=m.1,approx=a.0,diff=d.1,bound=b.1)
df.2 = data.frame(p.2,mean=m.2,approx=a.0,diff=d.2,bound=b.2)
df.3 = data.frame(p.3,mean=m.3,approx=a.0,diff=d.3,bound=b.3)

show.results(df.0)
show.results(df.1)
show.results(df.2)
show.results(df.3)
```